---
title: "Case_study_1_Cyclistic"
author: "Miracle"
date: "2023-03-22"
output:
  pdf_document: default
  html_document: default
---

# Merging monthly reports into quarterly files

Let me introduce this case study analysis. We are given data about the bike-share company which includes It is believed that the company's future success depends on maximizing the number of annual memberships. Therefore, we want to understand how casual riders and annual members use bikes differently. From these insights, we will try to design a new marketing strategy to convert casual riders into annual members. Necessary data can be downloaded [here](https://divvy-tripdata.s3.amazonaws.com/index.html).

### STEP-1 We start with loading required tools and setting working directory

```{r, setup, include=FALSE}
knitr::opts_knit$set(root.dir = "C:/my filess/Data-Analysis/Case_study_1_Cyclistic_bike-share")
```

```{r}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
getwd() #displays your working directory

```

### STEP-2 Then we upload necessary documents from directory and name them

```{r}
march22_table<-read.csv("202203-divvy-tripdata.csv")
april22_table<-read.csv("202204-divvy-tripdata.csv")
may22_table<-read.csv("202205-divvy-tripdata.csv")
```

### STEP-3 Before merging them, we have to check if the names of columns are consistent

While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r}
colnames(march22_table)
colnames(april22_table)
colnames(may22_table)
```

### STEP-4 Now we change ride_id and rideable_type columns to character so they can stack correctly

```{r}
march22_table<- mutate(march22_table, ride_id=as.character(ride_id), 
                     rideable_type=as.character(rideable_type))
april22_table<- mutate(april22_table, ride_id=as.character(ride_id), 
                     rideable_type=as.character(rideable_type))
may22_table<- mutate(may22_table, ride_id=as.character(ride_id), 
                   rideable_type=as.character(rideable_type))

```

### STEP-5 Then we recheck the dataframes

```{r}
str(march22_table)
str(april22_table)
str(may22_table)
```

### STEP-6 Finally, we start to merge individual monthly files into quarterly reports

```{r}
spring_2022<- bind_rows(march22_table,april22_table,may22_table)
```

### STEP-6 Now it is time to inspect the created table

```{r}
colnames(spring_2022)   # list of column names
head(spring_2022)  # to see first 6 rows 
tail(spring_2022)  # to see last 6 rows
nrow(spring_2022)  # how many rows in the table
dim(spring_2022)   # Dimensions of the data
summary(spring_2022)  # Statistical summary of the data, mainly numeric
```

### (SUMMER) Then we will repeat given steps for other monthly reports too Uploading files

```{r}
june22_table<-read.csv("202206-divvy-tripdata.csv")
july22_table<-read.csv("202207-divvy-tripdata.csv")
august22_table<-read.csv("202208-divvy-tripdata.csv")
```

Check columns if they match perfectly

```{r}
colnames(june22_table)
colnames(july22_table)
colnames(august22_table)
```

Change to character of ride_id and rideable_type columns

```{r}
june22_table<- mutate(june22_table, ride_id=as.character(ride_id), 
                     rideable_type=as.character(rideable_type))
july22_table<- mutate(july22_table, ride_id=as.character(ride_id), 
                     rideable_type=as.character(rideable_type))
august22_table<- mutate(august22_table, ride_id=as.character(ride_id), 
                   rideable_type=as.character(rideable_type))

```

Checking dataframe for errors

```{r}
str(june22_table)
str(july22_table)
str(august22_table)
```

### Last step: Merging

```{r}
summer_2022<-bind_rows(june22_table,july22_table,august22_table)
```

Inspect the result

```{r}
colnames(summer_2022)   # list of column names
head(summer_2022)  # to see first 6 rows 
tail(summer_2022)  # to see last 6 rows
nrow(summer_2022)  # how many rows in the table
dim(summer_2022)   # Dimensions of the data
summary(summer_2022)  # Statistical summary of the data, mainly numeric
```

# Fall quarter

```{r}
september22_table<-read.csv("202209-divvy-tripdata.csv")
october22_table<-read.csv("202210-divvy-tripdata.csv")
november22_table<-read.csv("202211-divvy-tripdata.csv")
```

Check the column names

```{r}
colnames(september22_table)
colnames(october22_table)
colnames(november22_table)
```

Change to character

```{r}
september22_table<- mutate(september22_table, ride_id=as.character(ride_id), 
                     rideable_type=as.character(rideable_type))
october22_table<- mutate(october22_table, ride_id=as.character(ride_id), 
                     rideable_type=as.character(rideable_type))
november22_table<- mutate(november22_table, ride_id=as.character(ride_id), 
                   rideable_type=as.character(rideable_type))
```

Check the dataframes

```{r}
str(september22_table)
str(october22_table)
str(november22_table)
```

Merging

```{r}
fall_2022<-bind_rows(september22_table,october22_table,november22_table)
```

Inspecting the results

```{r}
colnames(fall_2022)   # list of column names
head(fall_2022)  # to see first 6 rows 
tail(fall_2022)  # to see last 6 rows
nrow(fall_2022)  # how many rows in the table
dim(fall_2022)   # Dimensions of the data
summary(fall_2022)  # Statistical summary of the data, mainly numeric
```

# Winter quarter

Upload the docs

```{r}
december22_table<-read.csv("202212-divvy-tripdata.csv")
january23_table<-read.csv("202301-divvy-tripdata.csv")
february23_table<-read.csv("202302-divvy-tripdata.csv")
```

Check the column names if they are the same

```{r}
colnames(december22_table)
colnames(january23_table)
colnames(february23_table)
```

Change dataframe of ride_id and rideable_type to character

```{r}
december22_table<-mutate(december22_table, ride_id=as.character(ride_id),
                         rideable_type=as.character(rideable_type))
january23_table<-mutate(january23_table, ride_id=as.character(ride_id),
                        rideable_type=as.character(rideable_type))
february23_table<-mutate(february23_table, ride_id=as.character(ride_id),
                         rideable_type=as.character(rideable_type))
```

Check the dataframe before merging

```{r}
str(december22_table)
str(january23_table)
str(february23_table)
```

Last step is merging

```{r}
winter_2023<- bind_rows(december22_table,january23_table,february23_table)
```

And we Inspect the returned data

```{r}
colnames(winter_2023)
head(winter_2023)
tail(winter_2023)
dim(winter_2023)
summary(winter_2023)
```

And it is done!

We can take it further from here and make the annual report too. For this, we just have to combine quarterly reports into yearly. Since we have checked everything before we just have to use 'bind_raws' function

```{r}
annual_report_22_23<-bind_rows(spring_2022,summer_2022,fall_2022,winter_2023)
```

# ============================ ANALYSIS ==========================

Before starting there are few problems we need to fix:

## (1) The data can only be aggregated at the ride-level, which is too granular. We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate and analyze the data.

## (2) We will want to add a calculated field for length of ride to calculate how long members vs. casual riders use the bikes. We will add "ride_length" to the entire dataframe for consistency.

## (3) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

## (4) We check if different labels are used for 'member_casual' column

```{r}
table(annual_report_22_23$member_casual)
```

Since names are consistent we continue to solve other problems before analyzing

## Add columns that list the date, month, day, and year of each ride

This will allow us to aggregate ride data for each month, day, or year ... before completing these operations we could only aggregate at the ride level

```{r}
annual_report_22_23$date<-as.Date(annual_report_22_23$started_at) # The default format is yyyy-mm-dd
annual_report_22_23$month<-format(as.Date(annual_report_22_23$date),"%m")
annual_report_22_23$day<-format(as.Date(annual_report_22_23$date),"%d")
annual_report_22_23$year<-format(as.Date(annual_report_22_23$date), "%Y")
annual_report_22_23$day_of_week<-format(as.Date(annual_report_22_23$date), "%A")
```

## Add a "ride_length" calculation to all_trips (in seconds)

We use 'difftime()' function to calculate the time difference between 'started_at' and 'ended_at'

```{r}
annual_report_22_23$ride_length<-difftime(annual_report_22_23$ended_at,annual_report_22_23$started_at)
```

Check the structure of the data to see the changes we made so far

```{r}
str(annual_report_22_23)
```

We found out that 'ride_length' is not numeric value but factor value so we will change it to numeric to perform calculations on it

```{r}
is.factor(annual_report_22_23$ride_length) #to check if it is factor
annual_report_22_23$ride_length<-as.numeric(annual_report_22_23$ride_length)
is.numeric(annual_report_22_23)   #to check if it changed

```

## Remove "bad" data

The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative We will create a new version of the dataframe (v2) since data is being removed

```{r}
annual_report_v2<-annual_report_22_23[!(annual_report_22_23$start_station_name== "HQ QR"|annual_report_22_23$ride_length<0),]
```

# ============================ DESCRIPTIVE ANALYSIS ==========================

## Descriptive analysis on ride_length (all figures in seconds)

```{r}
mean(annual_report_v2$ride_length) #straight average (total ride length / rides)
median(annual_report_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(annual_report_v2$ride_length) #longest ride
min(annual_report_v2$ride_length) #shortest ride

```

We can condense the four lines above to one line using summary() on the specific attribute

```{r}
summary(annual_report_v2$ride_length)
```

## Compare casual riders and members

```{r}
aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual, FUN = mean)
aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual, FUN = median)
aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual, FUN = max)
aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual, FUN = min)
```

## Let's check average ride time by each day for members and casual users

```{r}
aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual + 
            annual_report_v2$day_of_week, FUN = mean)
```

Now if we look at the results the days of week are out of order So we will fix it:

```{r}
annual_report_v2$day_of_week<-ordered(annual_report_v2$day_of_week, levels=c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))
```

We recheck the average ride time with ordered days of week

```{r}
aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual + 
            annual_report_v2$day_of_week, FUN = mean)
```

## It is time to analyze ridership data by type and weekday

```{r}
annual_report_v2 %>% 
  mutate(weekday = wday(started_at, label=TRUE)) %>% #this creates weekday field using 'wday()'
  group_by(member_casual,weekday) %>% #this one groups by usertype and weekday
  summarize(number_of_rides=n(),average_duration = mean(ride_length)) %>% #calculate average duration 
  arrange(member_casual,weekday) #sorts
```

## Let's visualize the number of rides by rider type

```{r}
annual_report_v2 %>% 
  mutate(weekday = wday(started_at, label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarize(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday) %>% 
  ggplot(aes(x=weekday,y=number_of_rides,fill=member_casual))+
  geom_col(position = "dodge")
```

## Next, let's create a visualization for average duration

```{r}
annual_report_v2 %>% 
  mutate(weekday = wday(started_at, label=TRUE)) %>% 
  group_by(member_casual,weekday) %>% 
  summarise(number_of_rides = n(), average_duration = mean(ride_length)) %>% 
  arrange(member_casual,weekday) %>% 
  ggplot(aes(x=weekday,y=average_duration,fill=member_casual))+
  geom_col(position = "dodge")
```

## Lastly, we will create a csv file that we can vizualize in Excel, Tableau or PowerPoint

We have to set the destination for this file to save.

```{r}
counts<-aggregate(annual_report_v2$ride_length~annual_report_v2$member_casual + 
                    annual_report_v2$day_of_week, FUN = mean)
write.csv(counts, file = "C:/my filess/Data-Analysis/Case_study_1_Cyclistic_Bike-share/average_ride_length.csv")
```

# =========DONE!=========